Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-03-18 code-tests.tex
corrected tex typo

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{One-zone collapse test}
\label{sec.tests.1-zone}
The one-zone collapse test simulates the collapse of a
self-gravitating gas cloud using a semi-analytic model for the
evolution of gas density and adiabatic heat input as a function of
time.  It is designed to test the chemistry and cooling modules over a
wide range in densities and over physically motivated timescales.
Because this test disables the hydrodynamic and gravity solvers and
uses a simple model for the density evolution, it is far faster than
running a true collapse simulation.  The density evolution is based on
the self-similar Larson-Penston \citep{1969MNRAS.145..271L,
  1969MNRAS.144..425P} solution for isothermal collapse with a
modification to account for the efficiency with which the heat
introduced by compression can be radiated away
\citep{1983ApJ...265.1047Y}.  Our implementation, described briefly
here, follows the work of \citet{2005ApJ...626..627O}, which one
should consult for further details.  The density evolution is given by 
\begin{equation}
\frac{d\rho}{dt} = \frac{\rho}{t_{col}},
\end{equation}
where the collapse timescale, $t_{col}$, is
\begin{equation} \label{eqn.tcol}
t_{col} = \frac{t_{dyn}}{\sqrt{1 - f}},
\end{equation}
and $t_{dyn}$ is the dynamical time, expressed as
\begin{equation}
t_{dyn} = \sqrt{\frac{3 \pi}{32 G \rho}}.
\end{equation}
The collapse timescale is altered from the dynamical time by the
factor, $f$ in Equation \ref{eqn.tcol}, which an approximation of the
ratio of the pressure to the force of gravity.  The value of $f$
depends on the effective adiabatic index, $\gamma_{ef} \equiv (\partial
ln\ p / \partial ln\ \rho)$, which we calculate by linearly extrapolating
from derivative values for the two previous timesteps.  For the value
of $f$, we use the piecewise function of \citet{2005ApJ...626..627O},
given by
\begin{equation}
f = \left\{
  \begin{array}{ll}
  0, & \gamma_{ef} < 0.83,\\
  0.6 + 2.5 (\gamma_{ef} - 1) - 6.0 (\gamma_{ef} - 1)^{2}, & 0.83 <
  \gamma_{ef} < 1,\\
  1.0 + 0.2 (\gamma_{ef} - 4/3) - 2.9 (\gamma_{ef} - 4/3)^{2}, & \gamma_{ef} > 1.}
\end{array} \right.
\end{equation}
The specific energy evolves as
\begin{equation}
\frac{de}{dt} = -p \frac{d}{dt} \frac{1}{\rho} - \Lambda,
\end{equation}
where $\Lambda$ is the cooling rate in units of erg s$^{-1}$ g$^{-1}$
and energy, temperature, density, and pressure are related by the
ideal gas law.  Figure \ref{fig.onezone} shows an example of the
one-zone collapse test performed with an initial number density of 1
cm$^{-3}$ and temperature of 100 K using the 12 species chemistry
network with H, D, and He species and metal cooling rates calculated
with Cloudy.

\begin{figure} \label{fig.onezone}
  \begin{center}
    \includegraphics[width=1.0\textwidth]{figures/OneZoneCollapseTest.eps}
    \caption{Evolution of temperature versus number density for a one-zone
      collapse test for gas at metallicities from 0 to 10$^{-2} Z_{\odot}$
      using the primordial chemistry network with tabulated metal cooling
      rates calculated with Cloudy.}
    \label{fig.1zone}
  \end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Photo-evaporation of a dense clump}
\label{sec.tests.raytracing}
\red{(John)}
Photo-evaporation of a dense clump.  Tests ray tracing, specifically
the shadowing effects of the clump, and the hydrodynamic response of
the clump.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{One-zone collapse test}
\label{sec.tests.1-zone}
The one-zone collapse test simulates the collapse of a
self-gravitating gas cloud using a semi-analytic model for the
evolution of gas density and adiabatic heat input as a function of
time.  It is designed to test the chemistry and cooling modules over a
wide range in densities and over physically motivated timescales.
Because this test disables the hydrodynamic and gravity solvers and
uses a simple model for the density evolution, it is far faster than
running a true collapse simulation.  The density evolution is based on
the self-similar Larson-Penston \citep{1969MNRAS.145..271L,
  1969MNRAS.144..425P} solution for isothermal collapse with a
modification to account for the efficiency with which the heat
introduced by compression can be radiated away
\citep{1983ApJ...265.1047Y}.  Our implementation, described briefly
here, follows the work of \citet{2005ApJ...626..627O}, which one
should consult for further details.  The density evolution is given by 
\begin{equation}
\frac{d\rho}{dt} = \frac{\rho}{t_{col}},
\end{equation}
where the collapse timescale, $t_{col}$, is
\begin{equation} \label{eqn.tcol}
t_{col} = \frac{t_{dyn}}{\sqrt{1 - f}},
\end{equation}
and $t_{dyn}$ is the dynamical time, expressed as
\begin{equation}
t_{dyn} = \sqrt{\frac{3 \pi}{32 G \rho}}.
\end{equation}
The collapse timescale is altered from the dynamical time by the
factor, $f$ in Equation \ref{eqn.tcol}, which an approximation of the
ratio of the pressure to the force of gravity.  The value of $f$
depends on the effective adiabatic index, $\gamma_{ef} \equiv (\partial
ln\ p / \partial ln\ \rho)$, which we calculate by linearly extrapolating
from derivative values for the two previous timesteps.  For the value
of $f$, we use the piecewise function of \citet{2005ApJ...626..627O},
given by
\begin{equation}
f = \left\{
  \begin{array}{ll}
  0, & \gamma_{ef} < 0.83,\\
  0.6 + 2.5 (\gamma_{ef} - 1) - 6.0 (\gamma_{ef} - 1)^{2}, & 0.83 <
  \gamma_{ef} < 1,\\
  1.0 + 0.2 (\gamma_{ef} - 4/3) - 2.9 (\gamma_{ef} - 4/3)^{2}, & \gamma_{ef} > 1.
\end{array} \right.
\end{equation}
The specific energy evolves as
\begin{equation}
\frac{de}{dt} = -p \frac{d}{dt} \frac{1}{\rho} - \Lambda,
\end{equation}
where $\Lambda$ is the cooling rate in units of erg s$^{-1}$ g$^{-1}$
and energy, temperature, density, and pressure are related by the
ideal gas law.  Figure \ref{fig.onezone} shows an example of the
one-zone collapse test performed with an initial number density of 1
cm$^{-3}$ and temperature of 100 K using the 12 species chemistry
network with H, D, and He species and metal cooling rates calculated
with Cloudy.

\begin{figure}
  \begin{center}
    \includegraphics[width=1.0\textwidth]{figures/OneZoneCollapseTest.eps}
  \end{center}
  \caption{Evolution of temperature versus number density for a one-zone
    collapse test for gas at metallicities from 0 to 10$^{-2} Z_{\odot}$
    using the primordial chemistry network with tabulated metal cooling
    rates calculated with Cloudy.}
  \label{fig.onezone}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Photo-evaporation of a dense clump}
\label{sec.tests.raytracing}

The photo-evaporation of dense clumps is prevalent in radiation
hydrodynamics simulations, and this test problem examines the
ionization front propagation into the dense clump, shadowing effects
behind the clump, and the hydrodynamic response on the clump from
photo-heating.  The problem setup is the same as Test 7 in the
Cosmological Radiative Transfer Comparison Project
\citep{IlievEtAl2009} and \citet{Wise11_Moray}.  The simulation domain
is 6.6~kpc on a side with an ambient medium of pure neutral hydrogen
$n_{\rm H} = 2 \times 10^{-4}\; \cubecm$ and $T = 8000 \unit{K}$.  We
place a spherical overdensity in hydrostatic equilibrium with the
ambient medium.  It has a radius $r = 0.8 \unit{kpc}$, hydrogen
density $n_{\rm H} = 0.04 \cubecm$, and temperature $T = 40$ K, and it
is centered at $(x,y,z) = (5, 3.3, 3.3) \unit{kpc}$.  In
\citet{IlievEtAl2009}, all of the codes used a fixed $128^3$ grid to
ease the comparison, but in this test to demonstrate a higher
resolution AMR solution, we employ a $128^3$ grid with two additional
levels of refinement with cells with a baryon mass (method 2 in
\S\ref{sec:refinement_criteria}) greater than 1.5 being flagged for
refinement.

This cloud is subject to radiation from a point source at the center
of the $y=0$ boundary with an ionizing photon luminosity
$\dot{N}_\gamma = 3 \times 10^{51}$ photons s$^{-1}$, corresponding to
a flux $F_0 = 10^6 \unit{photons s}^{-1} \unit{cm}^{-2}$ at the clump
surface closest to the radiation source.  The radiation source has a
spectrum of a $T = 10^5 \unit{K}$ blackbody, and we use four energy
groups with the following mean energies and relative luminosities:
$E_i = (17.98, 31.15, 49.09, 76.98) \unit{eV}, L_i/L = (0.23, 0.36,
0.24, 0.06)$ that are optimized to reduce errors in the solutions with
a full spectrum and energy discretization \citep{Mirocha12}.  Note
that this choice of energy groups is different from
\citet{Wise11_Moray}.

### 2013-03-18 overview.tex

\begin{eqnarray} 
\frac{\partial \rho}{\partial t} 
          + \frac{1}{a} \nabla \cdot (\rho \vecv) & = & 0
        \label{eq:mass} \\
%
\frac{\partial \rho \vecv}{\partial t}  
   + \frac{1}{a} \nabla \cdot \left(\rho \vecv \vecv + p^* - \vecB \vecB \right) & = &
    - \frac{\dot{a}}{a} \rho \vecv - \frac{1}{a} \rho \nabla \phi
    %
%          + \frac{1}{a} ( \vecv \cdot \nabla ) \vecv
%    & = & - \frac{\dot{a}}{a} \vecv 
%          - \frac{1}{a \rho} \nabla p
%          - \frac{1}{a \rho} (\nalba \times \vecB) \times \vecB
%          - \frac{1}{a} \nabla \phi, 
        \label{eq:momentum} \\
%
\frac{\partial E}{\partial t} 
+ \frac{1}{a} \nabla \cdot \left[ (E + p^*) \vecv -  \vecB (\vecB \cdot \vecv) \right] & = &
     - \frac{\dot{a}}{a} \left( 2E - \frac{\vecB^2}{2} \right)  - \frac{1}{a} \vecv \cdot \nabla \phi 
     %
%         + \frac{1}{a} \vecv \cdot \nabla E
 %  & = & - \frac{\dot{a}}{a} (3 \frac{p}{\rho} + {\vecv \cdot \vecv} )
  %       - \frac{1}{a \rho } \nabla \cdot (p \vecv) 
%%\nonumber   & &   %uncomment for 2-column
 %        - \frac{1}{a} \vecv \cdot \nabla \phi 
 %               + \Gamma - \Lambda.
       \label{eq:total_energy}  \\
%
\frac{\partial \vecB}{\partial t} - \frac{1}{a}  \nabla \times (\vecv \times \vecB) & = & 0 \label{eq:induction}
\end{eqnarray}

%\begin{eqnarray} 
%\frac{\partial \rho}{\partial t} 
%          + \vecv \cdot \nabla \rho
%    & = & - \rho \nabla \cdot \vecv,
%        \label{eq:mass} \\
%%
%\frac{\partial \vecv }{ \partial t} 
%          + ( \vecv \cdot \nabla ) \vecv 
%    & = & - \frac{1}{\rho} \nabla p
%          - \nabla \phi,
%        \label{eq:momentum} \\
%%
% \frac{\partial E}{\partial t} + \vecv \cdot \nabla E
%    & = & - \frac{1}{\rho} \nabla \cdot (p \vecv) 
%          - \vecv \cdot \nabla \phi
%%\nonumber          \\  & &     % uncomment for 2-column
%     + \Gamma - \Lambda
%        \label{eq:total_energy}
%\end{eqnarray}
%
Here $E$, $\rho$, $\vecv$, and $\vecB$ are the comoving total fluid energy, comoving baryonic density, peculiar velocity, comoving magnetic field strength, respectively.  
The first equation represents conservation of mass, the second, conservation of momentum, and the third, conservation of total fluid (kinetic plus thermal) energy.  They are respectively, the first, second and third moments of the Boltzmann equation. Terms representing radiative cooling ($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the energy equation (\ref{eq:total_energy}).

The total fluid energy ($E$) is given by:
\begin{equation}
E =  \rho e + \frac{1}{2} \vecv \cdot \vecv + \vecB \cdot \vecB
        \label{eq:total_energy_def}.
\end{equation}
where $e$ is the comoving specific thermal energy, and the total isotropic pressure $p^*$ is given by
\begin{equation}
p^* = p + \frac{1}{2} \vecB^2
\end{equation}
The equations are closed with the equation of state and Poisson's equation (we work in the Newtonian limit):
%
\begin{eqnarray}
%
e   & = & p / [(\gamma - 1) \rho ],
        \label{eq:eq_of_state} \\
%
\nabla^2 \phi & = & 4 \pi G \rho_{total}
%                   + 3 p_{total}/c^2) - \Lambda. 
        \label{eq:potential}
\end{eqnarray}
%
The equation of state is shown here for an ideal gas with a ratio of specific heats $\gamma$.
Equation (\ref{eq:potential}) includes contributions to the gravitational potential $\phi$ from the total mass density $\rho_{\rm total}$).
% as well as any radiation or massless neutrino background ($p_{total}^\prime$) or cosmological constant ($\Lambda_c$), if we are interested in the cosmological case.

@@

For completeness, we note here that as part of the usual transformation to comoving coordinates, we have defined the following comoving quantities (primes indicate variables defined in the usual, proper, coordinate system).

This set of coupled hydrodynamic plus gravity equations is the standard one that \enzo solves.  However, in the cosmological case it is useful to transform these to a coordinate system that is comoving with the expanding universe \citep{Peebles93} (but note that \enzo\ can also operate in a fixed coordinate system). Specifically, we define:
\begin{eqnarray}
\vecx & = & \vecr/a, \\
\vecv & = & a \hspace{0.5mm} d\vecx/dt = 
              \vecvp - \dot{a}\vecx, \\
\rho    & = & a^3 \rhop,   \\
p       & = & a^3 \pp, \\
E       & = & \Ep - 
              \dot{a} \vecx \cdot \vecvp - 
              \frac{1}{2} \dot{a}^2 \vecx^2, \\
\phi    & = & \phip + \frac{1}{2} a \ddot{a} \vecx^2. \\
\vecB & = & a^2 \vecB^\prime
\end{eqnarray}
where primes indicate the usual proper quantities.

@@

\subsubsection{Magnetohydrodynamics}

\subsubsection{Radiation transport: ray-casting}

>>>>

\begin{eqnarray} 
\frac{\partial \rho}{\partial t} 
          + \frac{1}{a} \nabla \cdot (\rho \vecv) & = & 0
        \label{eq:mass} \\
%
\frac{\partial \rho \vecv}{\partial t}  
   + \frac{1}{a} \nabla \cdot \left(\rho \vecv \vecv + \myvec{I}p^* - \frac{\vecB \vecB}{a} \right) & = &
    - \frac{\dot{a}}{a} \rho \vecv - \frac{1}{a} \rho \nabla \phi
    %
%          + \frac{1}{a} ( \vecv \cdot \nabla ) \vecv
%    & = & - \frac{\dot{a}}{a} \vecv 
%          - \frac{1}{a \rho} \nabla p
%          - \frac{1}{a \rho} (\nalba \times \vecB) \times \vecB
%          - \frac{1}{a} \nabla \phi, 
        \label{eq:momentum} \\
%
\frac{\partial E}{\partial t} 
+ \frac{1}{a} \nabla \cdot \left[ (E + p^*) \vecv -  \frac{1}{a} \vecB (\vecB \cdot \vecv) \right] & = &
     - \frac{\dot{a}}{a} \left( 2E - \frac{B^2}{2a} \right)  - \frac{1}{a} \vecv \cdot \nabla \phi 
     %
%         + \frac{1}{a} \vecv \cdot \nabla E
 %  & = & - \frac{\dot{a}}{a} (3 \frac{p}{\rho} + {\vecv \cdot \vecv} )
  %       - \frac{1}{a \rho } \nabla \cdot (p \vecv) 
%%\nonumber   & &   %uncomment for 2-column
 %        - \frac{1}{a} \vecv \cdot \nabla \phi 
 %               + \Gamma - \Lambda.
       \label{eq:total_energy}  \\
%
\frac{\partial \vecB}{\partial t} - \frac{1}{a}  \nabla \times (\vecv \times \vecB) & = & 0 \label{eq:induction}
\end{eqnarray}

%\begin{eqnarray} 
%\frac{\partial \rho}{\partial t} 
%          + \vecv \cdot \nabla \rho
%    & = & - \rho \nabla \cdot \vecv,
%        \label{eq:mass} \\
%%
%\frac{\partial \vecv }{ \partial t} 
%          + ( \vecv \cdot \nabla ) \vecv 
%    & = & - \frac{1}{\rho} \nabla p
%          - \nabla \phi,
%        \label{eq:momentum} \\
%%
% \frac{\partial E}{\partial t} + \vecv \cdot \nabla E
%    & = & - \frac{1}{\rho} \nabla \cdot (p \vecv) 
%          - \vecv \cdot \nabla \phi
%%\nonumber          \\  & &     % uncomment for 2-column
%     + \Gamma - \Lambda
%        \label{eq:total_energy}
%\end{eqnarray}
%
Here $E$, $\rho$, $\vecv$, and $\vecB$ are the comoving total fluid energy, comoving baryonic density, peculiar velocity, and comoving magnetic field strength, respectively.  $\myvec{I}$ is the identity matrix.
The first equation represents conservation of mass, the second, conservation of momentum, and the third, conservation of total fluid (kinetic plus thermal) energy.  They are respectively, the first, second and third moments of the Boltzmann equation.  The fourth equation is the magnetic induction equation.   Terms representing radiative cooling ($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the energy equation (\ref{eq:total_energy}).

The total fluid energy $E$ is given by:
\begin{equation}
E =  \rho e + \frac{\rho v^2}{2}  + \frac{B^2}{2a}
        \label{eq:total_energy_def}.
\end{equation}
where $e$ is the comoving specific thermal energy, and the total isotropic pressure $p^*$ is given by
\begin{equation}
p^* = p + \frac{B^2}{2a}
\end{equation}
The equations are closed with the equation of state and Poisson's equation (we work in the Newtonian limit):
%
\begin{eqnarray}
%
e   & = & p / [(\gamma - 1) \rho ],
        \label{eq:eq_of_state} \\
%
\nabla^2 \phi & = & 4 \pi G \rho_{total}
%                   + 3 p_{total}/c^2) - \Lambda. 
        \label{eq:potential}
\end{eqnarray}
%
The equation of state is shown here for an ideal gas with a ratio of specific heats $\gamma$.
Equation (\ref{eq:potential}) includes contributions to the gravitational potential $\phi$ from the total mass density $\rho_{\rm total}$.  Although we write the equations including both MHD and comoving coordinates, the code can operate both in the pure hydrodynamic limit, and without cosmological expansion.

@@

For completeness, we note here that as part of the usual transformation to comoving coordinates, we have defined a set of comoving quantities.  See, for example, \citet{Peebles93} for more details.  Specifically, we define:
\begin{eqnarray}
\vecx & = & \vecr/a, \\
\vecv & = & a \hspace{0.5mm} d\vecx/dt = 
              \vecvp - \dot{a}\vecx, \\
\rho    & = & a^3 \rhop,   \\
p       & = & a^3 \pp, \\
E       & = & a^3 (\Ep - 
              \dot{a} \vecx \cdot \vecvp - 
              \frac{1}{2} \dot{a}^2 \vecx^2), \\
\phi    & = & \phip + \frac{1}{2} a \ddot{a} \vecx^2. \\
\vecB & = & a^2 \vecB^\prime
\end{eqnarray}
where primes indicate the usual quantities defined in a fixed frame frame, $\vecr$ is the proper distance and $\vecx$ is the comoving distance.

@@

%\subsubsection{Magnetohydrodynamics}

\subsubsection{Radiation transport: ray-tracing}
